### 2021-12-03 doc/submissions/2021_11_18_Nature/Dow_manuscript_Nature.Rmd
move folder

In recent decades, Earth's forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@friedlingstein_global_2021; @harris_global_2021].
A substantive portion of this CO~2~ sink occurs in temperate deciduous forests, which sequester >300 Tg C yr^-1^ [@pugh_role_2019]. <!-- See issue # 92: https://github.com/EcoClimLab/growth_phenology/issues/92. ALSO: The temperate deciduous sink is >300 Tg C yr-1, but Pugh et al. 2019 estimate that ~200 Tg C yr-1 is attributable to demography (forest regrowth), while >100 Tg C yr-1 is controlled by environmental drivers. The latter portion seems more likely to be sensitive to spring temperatures, and I originally cited this number, but upon further thought it's probably complex enough that that's a false distinction. --> 
The future behavior of this CO~2~ sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].

@@

Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum daily growth rates ($g_{max}$) and the DOY on which it occurred ($DOY_{g_{max}}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- (69 trees at SCBI and 138 trees at Harvard Forest) and diffuse-porous (54 trees at SCBI and 202 trees at Harvard Forest) species, which differ in growth phenology[@dorangeville_peak_2021]. These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).

@@

<!--Weather data -->
For the dendrometer band and leaf phenology analyses, climate data corresponding to the measurement periods were obtained from local weather stations at each focal site. 
For SCBI, weather data were obtained from a meteorological tower adjacent to the ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020]. 
The R package *climpact* (see www.climpact-sci.org)[@benestad_empiricalstatistical_2008] was used to plot temperatures for visual inspection and to identify readings that were >3 standard deviations away from yearly means, which were labeled as outliers and removed from the dataset. 
Gaps in the SCBI meteorological tower data were subsequently filled using temperature readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail). 
Daily temperature records for Harvard Forest, which had already been gap-filled based on other local records, were obtained from the Harvard Forest weather station[@boose_shaler_2021; @boose_fisher_2021]. 
For each site, we used records of daily maximum ($T_{max}$) and minimum temperatures ($T_{min}$).

@@

To rule out the possibility that observed patterns were strongly influenced by summer drought, we examined the relationship between spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010].
The latter was obtained from the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020; @begueria_standardized_2014;@vicente-serrano_multiscalar_2010a]. 
Linear models were run with 4-, 6-, and 12-month SPEI values of June, July, and August vs April $T_{max}$ to determine if warm spring temperatures were associated with greater summer drought stress in our data set. 
No significant correlations were found (all p>0.05).

@@

<!--QUILT PLOT METHODS-->
For the analysis of correlation between RWI and climate variables, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
Correlations between monthly climate and $RWI$ were assessed using *dplR*[@bunn_dendrochronology_2008] and *bootRes*[@zang_dendroclimatic_2013] in R v 4.0 (R Core Team, 2020), which correlated functions and bootstrapped confidence intervals for these relationships[@biondi_dendroclim2002_2004].
We analyzed these correlations for January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
To test for potential lag effects of spring temperatures on growth the following year, we also ran a version of the analysis extending back to include climate of every month of the previous year (`r EDF_tree_rings_TMAX_extended`).
Correlations and significance levels for months April-August are given in `r chron_table`.

<!--MULTIVARIATE ANALYSIS METHODS-->
We used a multivariate model to test for joint effects of April and summer $T_{max}$ on RWI.
We began by testing univariate correlations of $T_{max}$ over three summer windows: June, June-July, and May-August. 
Having determined that, among these, June-July explained the most variation, we then analyzed the joint effects of April $T_{max}$ and June-July $T_{max}$ on RWI for each chronology independently using the base lm() function in R. 
Slopes and p-values for each chronology are given in `r chron_table`. 
Although some models may have benefited from data transformations, we determined that assumptions of normality and homoscedasticity were sufficiently met for the purposes of this analysis.

# Acknowledgements {-#Acknowledgements} 
We gratefully acknowledge all researchers who assisted with data collection in the field and laboratory, particularly Tsun Fung Au, Joshua Bregy, James Dickens, Karen Heeter, Anna Hennage, Daniel King, James McGee, Benjamin Lockwood, Jennifer McGarvey, Victoria Meakem, Josh Oliver, Jessica Shue, Karly Schmidt-Simard, Brandon Strange, Alyssa Terrell, Brynn Taylor, Michael Thornton, Senna Robeson, Matt Wenzel, and Luke Wylie. 
Thanks to David A. Orwig and members of the ForestGEO Ecosystems & Climate Lab at SCBI for helpful feedback.
The research was funded by ForestGEO (Smithsonian). 
Collection of tree-ring samples was funded by a USDA Agriculture and Food Research Initiative grant 2017-67013-26191 and from the Indiana University Vice Provost for Research Faculty Research Program.

@@

# Data Availability Statement {-#Data} 
The datasets generated and analysed during the current study are available via GitHub in the *growth_phenology* repository of the ForestGEO Ecosystems & Climate Lab @ SCBI, (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI *[TBD]*).
Master versions of the dendrometer band data are available for SCBI via GitHub in the *Dendrobands* repository of the Smithsonian Conservation Biology Institute ForestGEO plot (https://github.com/SCBI-ForestGEO/Dendrobands), which is archived in Zenodo (DOI 10.5281/zenodo.5551143), and for Harvard Forest via the Harvard Forest Data Archive (https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF149). 
Weather data for SCBI were obtained from the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/Met_Stations/SCBI), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215), and the National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail), downloaded on [DATE].
Weather data for Harvard Forest are available through the Harvard Forest Data Archive (https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF001 AND https://harvardforest1.fas.harvard.edu/exist/apps/datasets/showData.html?id=HF000).
Climate data were obtained from CRU v.4.04 via the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/CRU), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215).
The Standardised Precipitation-Evapotranspiration Index was obtained from the ForestGEO Climate Data Portal v1.0 (https://github.com/forestgeo/Climate/tree/master/Climate_Data/SPEI), which is archived in Zenodo (DOI: 10.5281/zenodo.3958215).
Canopy foliage phenology data were extracted from the MCD12Q2 V6 Land Cover Dynamics product (a.k.a. MODIS Global Vegetation Phenology product) via Google Earth Engine (https://developers.google.com/earth-engine/datasets/catalog/MODIS_006_MCD12Q2#description). 
Original tree cores are archived at the institutions of various members of the author team and will be made available upon reasonable request. 
<!-- do we need to add ITRDB??-->

# Code Availability Statement {-#Code}
Data were analysed in the open source statistical software R (version 4.0).
We used packages *climwin* v.1.2.3 (https://cran.r-project.org/web/packages/climwin/index.html),  *dplR* v.1.0.2, and *bootRes* v1.2.4, *rstanarm* v.2.21.1, and functions from *Rdendrom* (https://github.com/seanmcm/RDendrom/).
We used *climpact* software v.1.2.8 (see www.climpact-sci.org).
All custom code is available through the EcoClimlab GitHub repository (https://github.com/EcoClimLab/growth_phenology) and archived in Zenodo (DOI: *[TBD]*).

>>>>

In recent decades, Earth's forests have sequestered ~20% of anthropogenic CO~2~ emissions, thereby slowing the pace of atmospheric CO~2~ accumulation and climate change[@pan_large_2011; @friedlingstein_global_2020].
A large portion of this CO~2~ sink occurs in temperate deciduous forests, which sequester >300 Tg C yr^-1^ (>30% of the total forest C sink)[@pugh_role_2019]. <!-- The temperate deciduous sink is >300 Tg C yr-1, but Pugh et al. 2019 estimate that ~200 Tg C yr-1 is attributable to demography (forest regrowth), while >100 Tg C yr-1 is controlled by environmental drivers. The latter portion seems more likely to be sensitive to spring temperatures, and I originally cited this number, but upon further thought it's probably complex enough that that's a false distinction. --> 
The future behavior of this CO~2~ sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020; @friedlingstein_global_2020].

@@

Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum daily growth rates ($g_{max}$) and the DOY on which it occurred ($DOY_{g_{max}}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@dorangeville_peak_2021]. These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).

@@

<!--Climate -->
For the dendrometer band and leaf phenology analyses, climate data corresponding to the measurement periods were obtained from local weather stations at each focal site. 
For SCBI, weather data were obtained from a meteorological tower adjacent to the ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020]. 
The R package *climpact* (see www.climpact-sci.org)[@benestad_empiricalstatistical_2008] was used to plot temperatures for visual inspection and to identify readings that were >3 standard deviations away from yearly means, which were labeled as outliers and removed from the dataset. 
Gaps in the SCBI meteorological tower data were subsequently filled using temperature readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail). 
Daily temperature records for Harvard Forest, which had already been gap-filled based on other local records, were obtained from the Harvard Forest weather station[@boose_shaler_2021; @boose_fisher_2021]. 
For each site, we used records of daily maximum ($T_{max}$) and minimum temperatures ($T_{min}$).

@@

To rule out the possibility that observed patterns were strongly influenced by summer drought, we examined the relationship between spring tempreatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010].
The latter was obtained from the ForestGEO Climate Data Portal v1.0 (https://forestgeo.github.io/Climate/)[@anderson-teixeira_forestgeo_2020; @begueria_standardized_2014;@vicente-serrano_multiscalar_2010a]. 
Linear models were run with 4-, 6-, and 12-month SPEI values of June, July, and August vs April $T_{max}$ to determine if warm spring temperatures lead to greater summer drought stress. 
No significant correlations were found (all p>0.05).

@@

<!--QUILT PLOT METHODS-->
For the analysis of correlation between RWI and climate variables, we obtained monthly $T_{max}$ and $T_{min}$ data for 1901-2019 from CRU v.4.04 [@harris_version_2020].
Correlations between monthly climate and $RWI$ were assessed using ‘dplR’[@bunn_dendrochronology_2008] and ‘bootRes’[@zang_dendroclimatic_2013] in R v 4.0 (R Core Team, 2020), which correlated functions and bootstrapped confidence intervals for these relationships[@biondi_dendroclim2002_2004].
We analyzed these correlations for January through September of the current year (presented in Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
To test for potential lag effects of spring temperatures on growth the following year, we also ran a version of the analysis extending back to include climate of every month of the previous year (`r EDF_tree_rings_TMAX_extended`).
Correlations and significance levels for months April-August are given in `r chron_table`.

<!--MULTIVARIATE ANALYSIS METHODS-->
We used a multivariate model to test for joint effects of April and summer $T_{max}$ on RWI.
We began by testing univariate correlations of $T_{max}$ over three summer windows: June, June-July, and May-August. 
Having determined that, among these, June-July explained the most variation, we then analyzed the joint effects of April $T_{max}$ and June-July $T_{max}$ on RWI for each chronology independently using the base lm() function in R. 
Slopes and p-values for each chronology are given in `r chron_table`.

# Acknowledgements {-#Acknowledgements} 
We gratefully acknowledge all researchers who assisted with data collection in the field and laboratory, particularly Tsun Fung Au, Joshua Bregy, James Dickens, Karen Heeter, Anna Hennage, Daniel King, James McGee, Benjamin Lockwood, Jennifer McGarvey, Victoria Meakem, Josh Oliver, Jessica Shue, Karly Schmidt-Simard, Brandon Strange, Alyssa Terrell, Brynn Taylor, Michael Thornton, Senna Robeson, Matt Wenzel, and Luke Wylie. 
Thanks to David A. Orwig and members of the ForestGEO Ecosystesms & Climate Lab at SCBI for helpful feedback.
The research was funded by ForestGEO (Smithsonian). 
Collection of tree-ring samples was funded by a USDA Agriculture and Food Research Initiative grant 2017-67013-26191 and from the Indiana University Vice Provost for Research Faculty Research Program.

@@
